---
title: "Testes Enrichment Analyses"
---
# Prep
## Specify paths
```{r paths}
#Input Files Root Path
path <- "path/to/pipelineInput/"
#Plots/Tables Output Path
plotsPath <- "path/to/plots/"
```

## Load Libraries
```{r libraries}
library(reshape2)
library(Rsamtools)
library(plyr)
library(sigfit)
library(devtools)
library(patchwork)
library(egg)
library(MutationalPatterns)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
ref_genome <- "BSgenome.Hsapiens.UCSC.hg19"
library(ref_genome, character.only = TRUE)
library(lme4)
library(tidyverse)
library(gtools)
select <- dplyr::select
rename <- dplyr::rename
mutate <- dplyr::mutate
slice <- dplyr::slice
options(scipen = 999)

```

# Functions
Functions to generate variants and callable regions for each dataset. These are run in the following section.

## Load Variant VCFs
```{r vcfLoadFunction}

load_VCF <- function(dataset, surveyed = NA) {
  #Pop variants from gnomAD
  if(dataset == "PopVars") {
    gnomadVCF <- read_vcfs_as_granges(paste0(path,"variantVCFs/PopVars/gnomad.genomes.r2.1.1.snps.filtered.singletons.vcf"), "gnomad", ref_genome, check_alleles=T)
    return(gnomadVCF)
  }
  #Testes by individual
  if(dataset == "Testes") {
    testesIndivFiles <- list.files(paste0(path, "variantVCFs/TestesNormal"), pattern = ".vcf.gz", full.names = TRUE)
    testesIndivSamples <-  str_remove(list.files(paste0(path, "variantVCFs/TestesNormal"), pattern = ".vcf.gz"), ".vcf.gz")
    testesIndivVCF <- read_vcfs_as_granges(testesIndivFiles, testesIndivSamples, ref_genome, check_alleles=T)
    return(testesIndivVCF)
  }
  #Colon by individual
  if(dataset == "Colon") {
    colonIndivFiles <- list.files(paste0(path, "variantVCFs/Colon"), pattern = ".vcf.gz", full.names = TRUE)
    colonIndivSamples <-  str_remove(list.files(paste0(path, "variantVCFs/Colon"), pattern = ".vcf.gz"), ".vcf.gz")
    colonIndivVCF <- read_vcfs_as_granges(colonIndivFiles, colonIndivSamples, ref_genome, check_alleles=T)
    return(colonIndivVCF)
  }
  #Trios from DeCode
  if(dataset == "Trios") {
    triosVCF <- read_vcfs_as_granges(paste0(path, "variantVCFs/Trios/trios_decode.vcf"), "Trio_data", ref_genome, check_alleles=T)
    return(triosVCF)
  }
  #PanBody Patient muts
  if(dataset == "PanBody") {
    panBodyFiles <- list.files(paste0(path, "variantVCFs/PanBody"), pattern = ".vcf.gz" , full.names = TRUE)
    panBodySamples <- str_remove(list.files(paste0(path, "variantVCFs/PanBody"), pattern = ".vcf.gz"), ".vcf.gz")
    panBodyVCF <- read_vcfs_as_granges(panBodyFiles, panBodySamples, ref_genome, check_alleles=T)
    return(panBodyVCF)
  }
}

```

## Load Surveyed Regions
*** Must be run as preloaded = F the first time, then preloaded = T can be used for subsequent runs with the .RData objects as shortcuts ***
```{r surveyedFunction}

load_callable <- function(dataset, preloaded = T) {
  #Full genome for hg19, no adjustment for callable regions
  if(dataset == "generic") {
    full_surveyed <- import(paste0(path, "surveyedRegions/full.callable.bed"))
    seqlevelsStyle(full_surveyed) <- "UCSC"
    return(full_surveyed)
  }
  #gnomad callable regions
  if(dataset == "PopVars") {
    pop_surveyed <- import(paste0(path, "surveyedRegions/PopVars/gnomadGenomesCallable.2.1.1.bed"))
    seqlevelsStyle(pop_surveyed) <- "UCSC"
    return(pop_surveyed)
  }
  #Testes: Genome coverage calculated by merging callable region bed files of each individual's samples
  if(dataset == "Testes") {
    if(preloaded) {
      load(paste0(path, "surveyedRegions/TestesNormal/testesIndivSurveyedList.RData"))
      return(testesIndivSurveyedList)
    } 
    else {
      testesIndivSamples <- list.files(paste0(path, "surveyedRegions/TestesNormal/"), pattern = ".sorted.merged.bed")
      testesIndivSurveyedList <- vector(mode = "list")
      for(i in testesIndivSamples){
        print(paste("Reading", i))
        surveyed <- import(paste0(path, "surveyedRegions/TestesNormal/",i))
        ## Import the file using rtracklayer and use the UCSC naming standard
        seqlevelsStyle(surveyed) <- "UCSC"
        testesIndivSurveyedList[[i]] <- surveyed
      }
      save(testesIndivSurveyedList, file= paste0(path,"surveyedRegions/TestesNormal/testesIndivSurveyedList.RData"))
      return(testesIndivSurveyedList)
    }
  }
  #Colon: Genome coverage calculated by merging callable region bed files of each individual's samples
  if(dataset == "Colon") {
    if(preloaded) {
      load(paste0(path,"surveyedRegions/Colon/colonIndivSurveyedList.RData"))
      return(colonIndivSurveyedList)
    } 
    else {
      colonIndivSamples <- list.files(paste0(path, "surveyedRegions/Colon/"), pattern = ".sorted.merged.bed")
      colonIndivSurveyedList <- vector(mode = "list")
      for(i in colonIndivSamples){
        print(paste("Reading", i))
        surveyed <- import(paste0(path, "surveyedRegions/Colon/",i))
        ## Import the file using rtracklayer and use the UCSC naming standard
        seqlevelsStyle(surveyed) <- "UCSC"
        colonIndivSurveyedList[[i]] <- surveyed
      }
      save(colonIndivSurveyedList, file=paste0(path,"surveyedRegions/Colon/colonIndivSurveyedList.RData"))
      return(colonIndivSurveyedList)
    }
  }
  #PanBody: Callable regions union of samples per indiv per tissue
  if(dataset == "PanBody") {
    if(preloaded) {
      load(paste0(path,"surveyedRegions/PanBody/PanBodySurveyedList.RData"))
      return(PanBodySurveyedList)
    } 
    else {
      PanBodySamples <- list.files(paste0(path, "surveyedRegions/PanBody/"), pattern = ".sorted.merged.bed")
      PanBodySurveyedList <- vector(mode = "list")
      for(i in PanBodySamples){
        print(paste("Reading", i))
        surveyed <- import(paste0(path, "surveyedRegions/PanBody/",i))
        surveyed <- keepStandardChromosomes(surveyed, pruning.mode = "coarse")
        surveyed <- dropSeqlevels(surveyed, "chrM", pruning.mode="coarse")
        ## Import the file using rtracklayer and use the UCSC naming standard
        seqlevelsStyle(surveyed) <- "UCSC"
        PanBodySurveyedList[[i]] <- surveyed
      }
      save(PanBodySurveyedList, file=paste0(path,"surveyedRegions/PanBody/PanBodySurveyedList.RData"))
      return(PanBodySurveyedList)
    }
  }
}
```

## Load TriRate List
```{r triRateCalc}

loadTriRate <- function(vcf_list, surveyed_list, tissues) {
  triRateDF <- tibble()
  for(tissue in unique(tissues)) {
    #Subset vcfs by tissue
    vcfsT <- vcf_list[tissues == tissue]
    #Get total muts for tissue
    muts0 <- mut_matrix(vcfsT, ref_genome)
    muts <- as_tibble(muts0) %>% 
      mutate(mutType = rownames(muts0)) %>% 
      rowwise() %>%
      mutate(totMuts = sum(c_across(where(is.numeric)))) %>% 
      mutate(context = paste0(str_sub(mutType,1,1),str_sub(mutType,3,3),str_sub(mutType,7,7))) %>% 
      group_by(context) %>% 
      dplyr::summarise(count = sum(totMuts), .groups = "drop")
      
    #Get mean opportunities for tissue
    surveyedT <- surveyed_list[tissues == tissue]
    for(i in 1:length(surveyedT)) {
      sequences <- getSeq(BSgenome.Hsapiens.UCSC.hg19, surveyedT[[i]])
      #Count trinuc freq
      tri_freqs <- trinucleotideFrequency(sequences)
      complement = vector()
      complement["A"] = "T"; complement["C"] = "G"; complement["G"] = "C"; complement["T"] = "A";	
      tri_freqs_summary <- as_tibble(tri_freqs) %>% 
        dplyr::summarise(across(.cols = everything(), sum)) %>% 
        pivot_longer(cols = everything(), names_to = "context0", values_to = "count") %>% 
        #Convert to pyr reference
        mutate(context = ifelse(str_sub(context0,2,2) %in% c("C", "T"), context0, paste0(complement[str_sub(context0, 3,3)], complement[str_sub(context0, 2,2)], complement[str_sub(context0, 1,1)]))) %>% 
        group_by(context) %>% 
        dplyr::summarise(count = sum(count), .groups = "drop")
      colnames(tri_freqs_summary) <- c("context", paste0("opps_", names(surveyedT[i])))
      muts <- muts %>% 
        left_join(tri_freqs_summary, by = "context")
    }
    freqDF <- muts %>% 
      rowwise() %>%
      mutate(meanOpps = mean(c_across(cols = starts_with("opps_")))) %>% 
      mutate(rate = count/meanOpps) %>% 
      mutate(rate = rate/sum(lengths(vcfsT))) %>% 
      mutate(tissue = tissue) %>% 
      select(context, tissue, rate)
    triRateDF <- triRateDF %>% 
      bind_rows(freqDF)
  }
  return(triRateDF)
}
```

## Custom Enrich Fxns
```{r customEnrichFxn}
#Edit intersect_with_region and genomic_distribution fxns from mutational patterns package
#to calculate expected number of mutations while controlling for trinucleotide mutation rate of the tissue type
intersect_with_regionCustom = function(vcf, surveyed, region, regionName, triRateTable, tissueType, sample, addSigs = F, sigsAssign = NULL) {
    # Number of mutations in vcf file
    n_muts = length(vcf)
    # Number of base pairs that were surveyed
    surveyed_length = sum(as.numeric(width(surveyed)))
    # Check if chromosome names are the same in the objects
    if (seqlevelsStyle(vcf) != seqlevelsStyle(surveyed))
        stop(paste("The chromosome names (seqlevels) of the VCF and the",
                    "surveyed GRanges object do not match."))
    if (seqlevelsStyle(region) != seqlevelsStyle(surveyed))
        stop(paste("The chromosome names (seqlevels) of the surveyed and",
                    "the region GRanges object do not match."))
    # Intersect genomic region and surveyed region
    surveyed_region = intersect(surveyed, region, ignore.strand = TRUE)
    surveyed_region_length = sum(width(surveyed_region))
    
    # Find which mutations lie in surveyed genomic region
    overlap = findOverlaps(vcf, surveyed_region)
    muts_in_region = unique(as.data.frame(as.matrix(overlap))$queryHits)
    observed = length(muts_in_region)
    
    #Add base to start and end to get correct trinuc counts
    surveyed_region@ranges@start <- as.integer(surveyed_region@ranges@start - 1)
    surveyed_region@ranges@width <- as.integer(surveyed_region@ranges@width + 2)
    #Count trinuc freq
    cat(paste0("Extracting sequences of ", tissueType, " in ", regionName, "...\n"))
  	sequences = getSeq(BSgenome.Hsapiens.UCSC.hg19, surveyed_region)
    cat("Calculating trinucleotide frequencies...\n")
    tri_freqs <- trinucleotideFrequency(sequences)
    
    complement = vector()
    complement["A"] = "T"; complement["C"] = "G"; complement["G"] = "C"; complement["T"] = "A";	
    tri_freqs_summary <- as.tibble(tri_freqs) %>% 
      dplyr::summarise(across(.cols = everything(), sum)) %>% 
      pivot_longer(cols = everything(), names_to = "context0", values_to = "count") %>% 
      #Convert to pyr reference
      mutate(context = ifelse(str_sub(context0,2,2) %in% c("C", "T"), context0, paste0(complement[str_sub(context0, 3,3)], complement[str_sub(context0, 2,2)], complement[str_sub(context0, 1,1)]))) %>% 
      group_by(context) %>% 
      dplyr::summarise(count = sum(count), .groups = "drop") %>% 
      left_join((triRateTable %>% filter(tissue == tissueType)), by = "context") %>% 
      mutate(expected = count * (rate * n_muts))
    
    expected = sum(tri_freqs_summary$expected)

    res = data.frame(n_muts,
                    surveyed_length,
                    regionName,
                    surveyed_region_length,
                    expected,
                    observed,
                    tissueType,
                    sample)
    
    #Add signature breakdown if table provided
    if(addSigs & observed > 0) {
      mutsInRegion <- list(vcf[muts_in_region])
      names(mutsInRegion) <- sample
      sigSums <- get_variant_contexts(mutsInRegion, ref_genome) %>% 
        left_join(sigsAssign, by = c("chr", "pos", "REF", "sub", "trinuc", "sample")) %>% 
        dplyr::summarise(across(matches("SBS"), sum))
      res <- bind_cols(res, sigSums)
    }
    #In case no variants observed
    if(addSigs & observed == 0) {
      sigCols <- sigsAssign %>% 
        select(matches("SBS")) %>% 
        slice(1) 
      sigCols[sigCols > 0] = 0
      res <- bind_cols(res, sigCols)
    }
    return(res)
}

genomic_distributionCustom <- function (vcf_list, surveyed_list, region_list, triRateTable, tissues,  addSigs = F, sigsAssign = NULL) {
    if (length(vcf_list) != length(surveyed_list)) 
        stop("vcf_list and surveyed_list must have the same length")
    if (is.null(names(region_list))) 
        stop(paste("Please set the names of region_list using:", 
            "    names(region_list) <- c(\"regionA\", \"regionB\", ...)", 
            sep = "\n"))
    df = data.frame()
    for (j in 1:length(region_list)) {
        for (i in 1:length(vcf_list)) {
            res = intersect_with_regionCustom(vcf_list[[i]], surveyed_list[[i]], 
                region_list[[j]], regionName = names(region_list)[j], triRateTable, tissueType = tissues[i], sample = names(vcf_list[i]), addSigs, sigsAssign)
            df = rbind(df, res)
        }
    }
    df$region = factor(df$region, levels = names(region_list))
    return(df)
}

calculateCIs <- function(expectedDF, vcf_list, regions, tissueTypes, n_sim) {
  
  region_names = names(regions)
  ciDF <- expectedDF %>% 
    mutate(lowerCI = NA) %>% 
    mutate(upperCI = NA) %>% 
    mutate(value = NA)
    
  for(tissue in unique(tissueTypes)) {

    #Create merged VCF for tissue
    vcf <- unlist(vcf_list[tissueTypes == tissue])
    #Region by tissue
    for(region in region_names) {
      cat(paste0("Calculating CIs for ", tissue, " in ", region, "...\n"))
      overlapRegion <- regions[region == region_names][[1]]
      
      observed <- length(unique(as.data.frame(as.matrix(findOverlaps(vcf, overlapRegion)))$queryHits))
      
      #Skip CIs for pop_vars as its too big to calculate effieciently
      if(tissue == "pop_vars") {
        ciDF$lowerCI[ciDF$by == tissue & ciDF$region == region] <- observed
        ciDF$upperCI[ciDF$by == tissue & ciDF$region == region] <- observed
        ciDF$value[ciDF$by == tissue & ciDF$region == region] <- observed
      }
      else {
        observations <- replicate(n_sim, length(unique(as.data.frame(as.matrix(findOverlaps(vcf[sort(sample(length(vcf), size = length(vcf),replace = T))], overlapRegion)))$queryHits)))
        cis <- quantile(observations, probs = c(0.025, 0.975))
        
        ciDF$lowerCI[ciDF$by == tissue & ciDF$region == region] <- cis[1]
        ciDF$upperCI[ciDF$by == tissue & ciDF$region == region] <- cis[2]
        ciDF$value[ciDF$by == tissue & ciDF$region == region] <- observed
      }
    }
  }
  return(ciDF)
}

```

# Specify Datasets
Select ONE of the preset groupings (or create custom set):

Warning: Loading and analyzing the set with popVars requires ~200GB memory, without have been using ~20GB
Reducing number of simulations for CIs will also speed up analysis significantly
Setting useSigs to false will speed up if only interested in main enrichment analyses

## a) Main
```{r main, eval = FALSE}
#Variants
colonVCF <- load_VCF("Colon")
popVarsVCF <- load_VCF("PopVars")
testesVCF <- load_VCF("Testes")
triosVCF <- load_VCF("Trios")
panBodyVCF <- load_VCF("PanBody")

VCFs <- c(colonVCF, popVarsVCF, testesVCF, triosVCF, panBodyVCF)
panBodyTissues <- str_sub(str_remove(list.files(paste0(path, "variantVCFs/PanBody"), pattern = ".vcf.gz"), ".vcf.gz"), 9)
tissues <- c(rep("colon",length(colonVCF)), "pop_vars", rep("testis",length(testesVCF)), "trio_DNMs", panBodyTissues)

#Surveyed Regions
colonSurveyed <- load_callable("Colon")
popVarsSurveyed <- load_callable("PopVars")
testesSurveyed <- load_callable("Testes")
triosSurveyed <- load_callable("PopVars")
panBodySurveyed <- load_callable("PanBody")
surveyed_list <- c(colonSurveyed, popVarsSurveyed, testesSurveyed, triosSurveyed, panBodySurveyed)


#Corrected pvalue threshold by # of tissues
p_corrected <- (0.05/length(unique(tissues)))

#Number of simulations for CIs
n_sims = 10000

#Should calculate sigs
useSigs = T

```

## b) MainNoPop
```{r mainNoPop, eval = FALSE}
#Variants
colonVCF <- load_VCF("Colon")
testesVCF <- load_VCF("Testes")
triosVCF <- load_VCF("Trios")
panBodyVCF <- load_VCF("PanBody")

VCFs <- c(colonVCF, testesVCF, triosVCF, panBodyVCF)
panBodyTissues <- str_sub(str_remove(list.files(paste0(path, "variantVCFs/PanBody"), pattern = ".vcf.gz"), ".vcf.gz"), 9)
tissues <- c(rep("colon",length(colonVCF)), rep("testis",length(testesVCF)), "trio_DNMs", panBodyTissues)

#Surveyed Regions
colonSurveyed <- load_callable("Colon")
testesSurveyed <- load_callable("Testes")
triosSurveyed <- load_callable("PopVars")
panBodySurveyed <- load_callable("PanBody")
surveyed_list <- c(colonSurveyed, testesSurveyed, triosSurveyed, panBodySurveyed)

#Corrected pvalue threshold by # of tissues
p_corrected <- (0.05/length(unique(tissues)))

#Number of simulations for CIs
n_sims = 10000

#Should calculate sigs
useSigs = T
```


# Tri rate tables
```{r triRateTables}
#Tri Rate Tables
triRateTable <- loadTriRate(VCFs, surveyed_list, tissues)
write_tsv(triRateTable, paste0(plotsPath, "triRateTable.tsv"))
p_triRateTable <- triRateTable %>% 
  ggplot(aes(x = context, y = rate, color = tissue)) +
  geom_point() + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90), legend.title = element_blank()) +
  ylab("Mut Rate") +
  xlab("") 

png(paste0(plotsPath, 'triRate.png'), height=4, width=12, units = "in", res = 300)
p_triRateTable
dev.off()

pdf(paste0(plotsPath, 'triRate.pdf'), height=4, width=12)
p_triRateTable
dev.off()

```


# Assign Sigs
```{r Assign Sigs}
probabilityDF <- NULL
if(useSigs) {
  get_variant_contexts <- function (vcf_list, ref_genome) {
  df_joined <- tibble()
    for (i in 1:length(vcf_list)) {
        vcf <- vcf_list[[i]]
        df <- tibble(chr = str_remove(as.vector(seqnames(vcf)), "chr"),
                     pos = vcf@ranges@start,
                     REF = as.character(vcf@elementMetadata@listData$REF),
                     sub = mut_type(vcf),
                     trinuc = type_context(vcf, ref_genome)[[2]],
                     sample = names(vcf_list[i]))
        df_joined <- df_joined %>% 
          bind_rows(df)
    }
  return(df_joined)
  }
  #Variants
  testesDF <- get_variant_contexts(testesVCF, ref_genome)
  colonDF <- get_variant_contexts(colonVCF, ref_genome)
  panBodyDF <- get_variant_contexts(panBodyVCF, ref_genome)
  variantDF <- bind_rows(testesDF, colonDF, panBodyDF) %>% 
    distinct()
  
  panBodyKey <- read_tsv(paste0(path, "Panbody_key_10082020.txt"), col_types = 'ccddcdccccc') %>%
    filter(SampleID != "PD43850w_P53_STM_G9") %>% 
    mutate(header1 = str_c(DonorID, TissueType1, sep = "_"))  %>%
    select(sample = SampleID, header1)  
  
  #Sample sig assigment
  mean_assignment0 <- read_tsv(paste0(path, "sigAssign/Decompose_Solution_Activities.txt"), col_types = cols(Samples = "c", .default = "d")) %>% 
    rename(sample = Samples) %>% 
    filter(sample != "PD43850w_P53_STM_G9") %>% 
    #Individual key for panBody
    left_join(panBodyKey, by = "sample") %>% 
    mutate(header1 = if_else(str_detect(sample, "lo0"), str_sub(sample, 1, 8), header1)) %>% 
    #Fix testes sample names
    mutate(header1 = str_replace(header1, "PD40745c", "PD40745c_PD42564c")) %>%
    mutate(header1 = str_replace(header1, "PD40746c", "PD40746c_PD42568c")) %>% 
    #Fix colon sample names
    mutate(header1 = str_replace(header1, "PD42564b", "PD40745b_PD42564b")) %>% 
    mutate(header1 = str_replace(header1, "PD42568b", "PD40746b_PD42568b")) %>% 
    #Summarise to indiv/tissue
    group_by(header1) %>% 
    dplyr::summarise(across(matches("SBS"), sum), .groups = "drop") %>% 
    rename(sample = header1)
  
  #Convert to fractions
  mean_assignment <- mean_assignment0 %>% 
    mutate(sum = (select_if(mean_assignment0, is.numeric) %>% rowSums())) %>% 
    mutate(across(matches("SBS"), ~ .x/sum)) %>% 
    select(sample, matches("SBS"))
  
  #Sigs
  data("cosmic_signatures_v3")
  sigs <- colnames(mean_assignment0)[-1]
  sigsDF <- as_tibble(t(cosmic_signatures_v3)) %>% 
    select(all_of(sigs)) %>% 
    mutate(sub = paste0(str_sub(colnames(cosmic_signatures_v3), 2,2), ">", str_sub(colnames(cosmic_signatures_v3), 6,6))) %>% 
    mutate(trinuc = str_sub(colnames(cosmic_signatures_v3), 1,3)) %>% 
    pivot_longer(cols = matches("SBS"), names_to = "sig", values_to = "freq")
  
  #Calculate probability of each variant belonging to each signature based on fraction of signature in sample and frequency of mutation type with that signature
  probabilityDF0 <- variantDF %>% 
    left_join(mean_assignment, by = "sample") %>% 
    pivot_longer(cols = matches("SBS"), names_to = "sig", values_to = "assignment") %>% 
    left_join(sigsDF, by = c("sub", "trinuc", "sig")) %>% 
    mutate(prob = assignment * freq)
  
  probabilityDF <- probabilityDF0 %>% 
    left_join(probabilityDF0 %>% group_by(chr, pos, REF, sub, trinuc, sample) %>% dplyr::summarise(probSum = sum(prob), .groups = "drop"), by = c("chr", "pos", "REF", "sub", "trinuc", "sample")) %>% 
    mutate(probNorm = prob/probSum) %>% 
    select(chr, pos, REF, sub, trinuc, sample, sig, probNorm) %>% 
    pivot_wider(names_from = sig, values_from = probNorm)
}

```

# Mutational Patterns Analysis
## Transcriptional strand bias analysis
```{r transcriptionBias, fig.height=6, fig.width=10}
genes_hg19 <- genes(TxDb.Hsapiens.UCSC.hg19.knownGene)
cds_hg19 <- cds(TxDb.Hsapiens.UCSC.hg19.knownGene)
transcripts_hg19 <- transcripts(TxDb.Hsapiens.UCSC.hg19.knownGene)

mut_mat_s <- mut_matrix_stranded(VCFs, ref_genome, transcripts_hg19)
#By Tissue Type
strand_counts <- strand_occurrences(mut_mat_s, by=tissues)
#Poisson test for strand asymmetry significance
strand_bias <- strand_bias_test(strand_counts) %>% 
  #Multiple test correction
  mutate(significant = if_else(p_poisson < p_corrected, "*", "")) %>% 
  mutate(logRatio = log2(transcribed/untranscribed))

write_tsv(strand_bias, path = paste0(plotsPath, 'strand_bias.tsv'))

```

## Replication Timing
```{r}
## Method from https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5536223/pdf/emss-73438.pdf
# regions bed files
regions_files = list.files(paste0(path,"mutPatternsRegions/genomic_regions"), full.names = T)[3:5]

print(regions_files)

# region bed files
regions_list=list()
for(i in regions_files){
  region <- import(i)
  ## Import the file using rtracklayer and use the UCSC naming standard
  seqlevelsStyle(region) <- "UCSC"
  regions_list[[i]] <- region
}

names(regions_list) = c("Early", "Intermediate", "Late")

#Adjusted rep timing
mainRepDist <- genomic_distributionCustom(VCFs, surveyed_list, regions_list, triRateTable, tissues, addSigs = useSigs, probabilityDF)
write_tsv(mainRepDist, path = paste0(plotsPath, 'repTimingIndiv.tsv'))

repTimingAdjusted <- mainRepDist %>% 
  group_by(region, tissueType) %>% 
  dplyr::summarise(across(where(is.numeric),sum), .groups = "drop") %>% 
  mutate(by = tissueType) %>% 
  select(by, region, n_muts, surveyed_length, surveyed_region_length, observed, expected)
write_tsv(repTimingAdjusted, path = paste0(plotsPath, 'repTimingSummarized.tsv'))

repTimingCIs <- calculateCIs(repTimingAdjusted, VCFs, regions_list, tissues, n_sim = n_sims)
write_tsv(repTimingCIs, path = paste0(plotsPath, 'repTimingCIs.tsv'))
```

## Introns vs Exons
```{r intronsExonsBurden}
#Exons vs Introns vs Coding
exons_hg19 <- exonicParts(TxDb.Hsapiens.UCSC.hg19.knownGene)
introns_hg19 <- intronicParts(TxDb.Hsapiens.UCSC.hg19.knownGene)
fiveUTRs_hg19 <- unlist(fiveUTRsByTranscript(TxDb.Hsapiens.UCSC.hg19.knownGene))
threeUTRs_hg19 <- unlist(threeUTRsByTranscript(TxDb.Hsapiens.UCSC.hg19.knownGene))
transcripts_hg19 <- transcripts(TxDb.Hsapiens.UCSC.hg19.knownGene)

#Exlude UTR regions
exons <- setdiff(setdiff(setdiff(exons_hg19, introns_hg19, ignore.strand = T), fiveUTRs_hg19, ignore.strand = T),threeUTRs_hg19, ignore.strand = T)
introns <- setdiff(setdiff(setdiff(introns_hg19, exons_hg19, ignore.strand = T), fiveUTRs_hg19, ignore.strand = T),threeUTRs_hg19, ignore.strand = T)
#Shrink genome coordinates on each end so that grabbing trinucleotide context at edges of chromosomes doesn't throw error
genome <- GRanges(seqinfo(base::get(ref_genome)))
genome@ranges@start <- as.integer(genome@ranges@start +1)
genome@ranges@width <- as.integer(genome@ranges@width -2)
intergenic <- GenomicRanges::setdiff(genome, transcripts_hg19, ignore.strand = T)

sum(exons@ranges@width)
sum(introns@ranges@width)
sum(intergenic@ranges@width)

exonicRegions_list=list()
exonicRegions_list[[1]] <- exons
exonicRegions_list[[2]] <- introns
exonicRegions_list[[3]] <- intergenic
names(exonicRegions_list) <- c("exonic", "intronic", "intergenic")

exonicDist <- genomic_distributionCustom(VCFs, surveyed_list, exonicRegions_list, triRateTable, tissues, addSigs = useSigs, probabilityDF)
write_tsv(exonicDist, path = paste0(plotsPath, 'exonicIndiv.tsv'))

exonicAdjusted <- exonicDist %>% 
  group_by(region, tissueType) %>% 
  dplyr::summarise(across(where(is.numeric),sum), .groups = "drop") %>% 
  mutate(by = tissueType)
write_tsv(exonicAdjusted, path = paste0(plotsPath, 'exonicSummarized.tsv'))

exonicCIs <- calculateCIs(exonicAdjusted, VCFs, exonicRegions_list, tissues, n_sim = n_sims)
write_tsv(exonicCIs, path = paste0(plotsPath, 'exonicCIs.tsv'))
```

# Expr/Meth Enrich
## PanBody Expr
```{r PanBodyExpr, fig.height=5, fig.width=20}
biomart <- read_tsv(paste0(path, "expression/txCoordsHG19.txt"), col_types = 'ccciiciic', skip = 1, col_names = c("gene", "geneVersion", "chr", "start", "end", "txType", "txStart", "txEnd", "strand")) %>% 
  mutate(strand = if_else(strand == "1", "+", "-")) %>% 
  group_by(gene, chr, strand) %>% 
  dplyr::summarise(start = min(start), end = max(end), txStart = min(txStart), txEnd = max(txEnd), txType = paste(txType, collapse = ";"), .groups = "drop")

gtexMedian <- read_tsv(paste0(path, "expression/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_median_tpm.gct.gz"), skip = 2, col_types = cols(Name = "c", Description = "c", .default = "d")) %>%  
  mutate(gene = str_remove(Name, "\\..*")) %>% 
  #Join hg19 coords
  left_join(biomart, by = "gene") %>% 
  #Drop non-standard chromosomes
  filter(chr %in% c("1","2","3","4","5","6","7","8","9","10","11","12","13","14","15","16","17","18","19","20","21","22","X")) %>% 
  drop_na()

txPanBody <- gtexMedian %>% 
  mutate(length = txEnd - txStart) %>% 
  select(gene, chr, txStart, txEnd, strand, length,
         adrenal_gland = "Adrenal Gland", bronchus_epithelium = "Lung", colon = "Colon - Sigmoid", liver = "Liver", oesophagus = "Esophagus - Mucosa",
         pancreas = "Pancreas", prostate = "Prostate", skin = "Skin - Sun Exposed (Lower leg)", #skinNoSun = "Skin - Not Sun Exposed (Suprapubic)", 
         small_bowel = "Small Intestine - Terminal Ileum", stomach = "Stomach", testis = "Testis", thyroid = "Thyroid", trio_DNMs = "Testis", pop_vars = "Testis", 
         benignTestis = "Testis", PostPubT = "Testis", PrePubT = "Testis", PostPubO = "Testis") %>% 
  #Pivot for one row per gene tissue pair
  pivot_longer(cols = -c(gene, chr, txStart, txEnd, strand, length), names_to = "tissue", values_to = "tpm") %>% 
  #Bin expression on tpm
  mutate(bin = "Unexp") %>% 
  mutate(bin = if_else(tpm > 0 & tpm <= 10, "0-1", bin)) %>% 
  mutate(bin = if_else(tpm > 1 & tpm <= 10, "1-10", bin)) %>% 
  mutate(bin = if_else(tpm > 10, "10+", bin)) %>% 
  mutate(bin = fct_relevel(bin, c("Unexp", "0-1", "1-10", "10+")))

exprData <- c("testis","colon","adrenal_gland","bronchus_epithelium","liver","oesophagus","prostate","skin","small_bowel","thyroid","pancreas","stomach", "trio_DNMs", "pop_vars", 
              "benignTestis", "PostPubT", "PrePubT", "PostPubO")
tissuesWithExprData <- exprData[exprData %in% tissues]

#Create granges lists for the 5 expression bins per tissue+
exprEnrichDF <- tibble()
for(etissue in tissuesWithExprData) {
  expr_list=list()
  for(eBin in c("Unexp", "0-1", "1-10", "10+")){
    exprDF <- txPanBody %>% filter(bin == eBin & tissue == etissue)
    expr_gr <- GRanges(exprDF$chr, IRanges(start=exprDF$txStart,end=exprDF$txEnd), strand = exprDF$strand)
    seqlevelsStyle(expr_gr) <- "UCSC"
    expr_list[[eBin]] <- expr_gr
  }
  names(expr_list) = c("Unexp", "0-1", "1-10", "10+")
  enrich <- genomic_distributionCustom(VCFs[etissue == tissues], surveyed_list[etissue == tissues], expr_list, triRateTable, tissues[etissue == tissues], addSigs = useSigs, probabilityDF)
  
  #Summarise across individuals
  exprEnrichAdjusted <- enrich %>% 
    group_by(region, tissueType) %>%
    dplyr::summarise(across(where(is.numeric),sum), .groups = "drop") %>%
    mutate(by = tissueType)
  
  #Add CIs
  exprEnrichCIs <- calculateCIs(exprEnrichAdjusted, VCFs[etissue == tissues], expr_list, tissues[etissue == tissues], n_sim = n_sims)

  #Add tissue to master DF
  exprEnrichDF <- bind_rows(exprEnrichDF, exprEnrichCIs)
}

write_tsv(exprEnrichDF, path = paste0(plotsPath, 'exprSummarizedCIs.tsv'))
```

# Summary Figure
## Point
```{r SummaryFigure, fig.height=8, fig.width=10}
tissueOrder <- c("pop_vars", "trio_DNMs","testis","adrenal_gland","appendix","bile_ducts","bronchus_epithelium","colon","liver","oesophagus","pancreas","prostate","skin","small_bowel","stomach","thyroid")

pRep <- repTimingCIs %>% mutate(region = str_replace(region, "Intermediate", "Int")) %>% 
  mutate(by = fct_relevel(by, tissueOrder)) %>% 
  ggplot(aes(x= region, y = observed/expected, color = region, group =1)) +
  geom_hline(yintercept = 1, colour = "grey") +
  geom_line(color = "darkgrey", linetype = "dashed") +
  geom_errorbar(aes(ymin=lowerCI/expected, ymax=upperCI/expected), width=.2, color = "black") +
  geom_point(alpha = 0.75) +
  scale_color_manual(values = c("#fcae91", "#fb6a4a", "#cb181d")) +
  facet_grid(~by, switch = "x") +
  labs(colour = "Replication timing bin") +
  ylab("obs/exp") +
  theme_bw() +
  theme(#axis.text.x = element_text(size = 8, angle = 45, vjust = 0.5),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        strip.background = element_blank(), #remove background for facet labels
        strip.text = element_text(angle = 45, vjust = 0.5),
        strip.placement="outside",
        panel.spacing = unit(0, "lines"), #remove space between facets
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())
pRep
pExonic <- exonicCIs %>% filter(region != "coding") %>% 
  mutate(by = fct_relevel(by, tissueOrder)) %>% 
  mutate(region = fct_relevel(region, c("intergenic","intronic", "exonic"))) %>% 
  ggplot(aes(x= region, y = observed/expected, color = region)) +
  geom_hline(yintercept = 1, colour = "grey") +
  scale_color_manual(values=c("#cccccc","#969696","#525252")) +
  geom_errorbar(aes(ymin=lowerCI/expected, ymax=upperCI/expected), width=.1, color = "black") +
  geom_point(alpha = 0.75) +
  facet_grid(~by, switch = "x") +
  labs(colour = "Gene region") +
  ylab("obs/exp") +
  theme_bw() +
  theme(#axis.text.x = element_text(size = 8, angle = 45, vjust = 0.5),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_blank(),
        panel.spacing = unit(0, "lines"), #remove space between facets
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())
pExonic 

pExpr <- exprEnrichDF %>% 
  #Add dummy lines to replace tissues with no expression data
  #bind_rows(tibble(by = rep(c("appendix", "bile_ducts"), each = 4), region = rep(c("Unexp", "0-1", "1-10", "10+"), 2), observed = 100, expected = 1)) %>% 
  mutate(region = fct_relevel(region, c("Unexp", "0-1", "1-10", "10+"))) %>% 
  mutate(by = fct_relevel(by, tissueOrder)) %>% 
  ggplot(aes(x= region, y = observed/expected, color = region, group = 1)) +
  geom_hline(yintercept = 1, colour = "grey") +
  geom_errorbar(aes(ymin=lowerCI/expected, ymax=upperCI/expected), width=.2, color = "black") +
  geom_point(alpha = 0.75) +
  geom_line(color = "darkgrey", linetype = "dashed") +
  facet_grid(~by, switch = "x") +
  scale_color_manual(values = c("#bdd7e7", "#6baed6", "#3182bd", "#08519c")) +
  labs(colour = "Expression (Median TPM)") +
  ylab("obs/exp") +
  ylim(NA, 1.60) +
  theme_bw() +
  theme(#axis.text.x = element_text(size = 8, angle = 45, vjust = 0.5),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_blank(),
        panel.spacing = unit(0, "lines"), #remove space between facets
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())
pExpr

pTxBias <- strand_bias %>% 
  mutate(by = fct_relevel(group, tissueOrder)) %>% 
  ggplot(aes(x= type, y = logRatio, fill = type, label = significant)) +
  geom_hline(yintercept = 0, colour = "grey") +
  # scale_color_manual(values=c("darkgrey","black")) +
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_text(color= "black", size = 3, aes(vjust = ifelse(sign(logRatio) > 0, 0.5, 1))) +
  scale_fill_manual(values = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854','#ffd92f')) +
  facet_grid(~by, switch = "x") +
  labs(fill = "Mutation type") +
  ylab("log2(trans/untrans)") +
  theme_bw() +
  theme(#axis.text.x = element_text(size = 8, angle = 45, vjust = 0.5),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_blank(),
        panel.spacing = unit(0, "lines"), #remove space between facets
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())
pTxBias


ggarrange(pTxBias, pExonic, pExpr, pRep, nrow = 4) 

png(paste0(plotsPath, 'summaryFig.png'), height=8, width=12, units = "in", res = 300)
ggarrange(pTxBias, pExonic, pExpr, pRep, nrow = 4)
dev.off()

pdf(paste0(plotsPath, 'summaryFig.pdf'), height=8, width=12)
ggarrange(pTxBias, pExonic, pExpr, pRep, nrow = 4)
dev.off()

```

## Sigs
```{r SummaryFigureSigs, fig.height=8, fig.width=10}
if(useSigs){
  sigOrder <- c("SBS1","SBS2","SBS3","SBS5","SBS7a","SBS7b","SBS7c","SBS7d","SBS8","SBS12","SBS13","SBS16","SBS18","SBS44","SBS57")
  
  pRep <- repTimingAdjusted %>% 
    group_by(region, tissueType) %>% 
    dplyr::summarise(across(where(is.numeric),sum), .groups = "drop") %>% 
    mutate(by = tissueType) %>% 
    mutate(region = str_replace(region, "Intermediate", "Int")) %>% 
    pivot_longer(cols = matches("SBS"), names_to = "sig", values_to = "sigCount") %>% 
    mutate(sig = fct_relevel(sig, sigOrder)) %>%
    mutate(by = fct_relevel(by, tissueOrder)) %>% 
    ggplot(aes(x= region, y = sigCount/expected, fill = sig, group =1)) +
    geom_hline(yintercept = 1, colour = "grey") +
    geom_bar( stat = "identity") +
    facet_grid(~by, switch = "x") +
    ylab("obs/exp") +
    theme_bw() +
    theme(axis.text.x = element_text(size = 8, angle = 45, vjust = 0.5),
          legend.position = "none",
          # axis.text.x = element_blank(),
          # axis.ticks.x = element_blank(),
          axis.title.x = element_blank(),
          strip.background = element_blank(), #remove background for facet labels
          strip.text = element_text(angle = 45, vjust = 0.5),
          strip.placement="outside",
          panel.spacing = unit(0, "lines"), #remove space between facets
          panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  pRep
  pExonic <- exonicAdjusted %>%
    mutate(by = tissueType) %>% 
    pivot_longer(cols = matches("SBS"), names_to = "sig", values_to = "sigCount") %>% 
    mutate(by = fct_relevel(by, tissueOrder)) %>% 
    mutate(sig = fct_relevel(sig, sigOrder)) %>%
    mutate(region = fct_relevel(region, c("intergenic","intronic", "exonic"))) %>% 
    ggplot(aes(x= region, y = sigCount/expected, fill = sig, group =1)) +
    geom_hline(yintercept = 1, colour = "grey") +
    geom_bar( stat = "identity") +
    facet_grid(~by, switch = "x") +
    ylab("obs/exp") +
    theme_bw() +
    theme(axis.text.x = element_text(size = 8, angle = 45, vjust = 0.5),
          legend.position = "none",
          # axis.text.x = element_blank(),
          # axis.ticks.x = element_blank(),
          axis.title.x = element_blank(),
          strip.background = element_blank(),
          strip.text.x = element_blank(),
          panel.spacing = unit(0, "lines"), #remove space between facets
          panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  pExonic 
  
  pExpr <- exprEnrichCIs %>% 
    pivot_longer(cols = matches("SBS"), names_to = "sig", values_to = "sigCount") %>% 
    #Add dummy lines to replace tissues with no expression data
    bind_rows(tibble(by = rep(c("appendix", "bile_ducts"), each = 4), region = rep(c("Unexp", "0-1", "1-10", "10+"), 2), observed = 100, expected = 1)) %>% 
    mutate(sig = fct_relevel(sig, sigOrder)) %>%
    mutate(region = fct_relevel(region, c("Unexp", "0-1", "1-10", "10+"))) %>% 
    mutate(by = fct_relevel(by, tissueOrder)) %>% 
    ggplot(aes(x= region, y = sigCount/expected, fill = sig, group =1)) +
    geom_hline(yintercept = 1, colour = "grey") +
    geom_bar( stat = "identity") +
    facet_grid(~by, switch = "x") +
    labs(fill = "Sigs") +
    ylab("obs/exp") +
    ylim(NA, 1.60) +
    theme_bw() +
    theme(axis.text.x = element_text(size = 8, angle = 45, vjust = 0.5),
          # axis.text.x = element_blank(),
          # axis.ticks.x = element_blank(),
          axis.title.x = element_blank(),
          strip.background = element_blank(),
          strip.text.x = element_blank(),
          panel.spacing = unit(0, "lines"), #remove space between facets
          panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  pExpr
  
  
  
  ggarrange(pExonic, pExpr, pRep, nrow = 3) 
  
  png(paste0(plotsPath, 'summaryFigSigs.png'), height=8, width=12, units = "in", res = 300)
  ggarrange(pExonic, pExpr, pRep, nrow = 3) 
  dev.off()
  
  pdf(paste0(plotsPath, 'summaryFigSigs.pdf'), height=8, width=12)
  ggarrange(pExonic, pExpr, pRep, nrow = 3) 
  dev.off()
}
```

## Sigs Prop
```{r SummaryFigureSigsProp, fig.height=8, fig.width=10}
if(useSigs){
  pRep <- repTimingAdjusted %>% 
    group_by(region, tissueType) %>% 
    dplyr::summarise(across(where(is.numeric),sum), .groups = "drop") %>% 
    mutate(across(matches("SBS"), ~ .x/observed)) %>% 
    mutate(by = tissueType) %>% 
    mutate(region = str_replace(region, "Intermediate", "Int")) %>% 
    pivot_longer(cols = matches("SBS"), names_to = "sig", values_to = "sigCount") %>% 
    mutate(sig = fct_relevel(sig, sigOrder)) %>%
    mutate(by = fct_relevel(by, tissueOrder)) %>% 
    ggplot(aes(x= region, y = sigCount, fill = sig, group =1)) +
    geom_hline(yintercept = 1, colour = "grey") +
    geom_bar( stat = "identity") +
    facet_grid(~by, switch = "x") +
    ylab("obs/exp") +
    theme_bw() +
    theme(axis.text.x = element_text(size = 8, angle = 45, vjust = 0.5),
          legend.position = "none",
          # axis.text.x = element_blank(),
          # axis.ticks.x = element_blank(),
          axis.title.x = element_blank(),
          strip.background = element_blank(), #remove background for facet labels
          strip.text = element_text(angle = 45, vjust = 0.5),
          strip.placement="outside",
          panel.spacing = unit(0, "lines"), #remove space between facets
          panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  pRep
  pExonic <- exonicAdjusted %>%
    mutate(by = tissueType) %>% 
    mutate(across(matches("SBS"), ~ .x/observed)) %>% 
    pivot_longer(cols = matches("SBS"), names_to = "sig", values_to = "sigCount") %>% 
    mutate(by = fct_relevel(by, tissueOrder)) %>% 
    mutate(sig = fct_relevel(sig, sigOrder)) %>%
    mutate(region = fct_relevel(region, c("intergenic","intronic", "exonic"))) %>% 
    ggplot(aes(x= region, y = sigCount, fill = sig, group =1)) +
    geom_hline(yintercept = 1, colour = "grey") +
    geom_bar( stat = "identity") +
    facet_grid(~by, switch = "x") +
    ylab("obs/exp") +
    theme_bw() +
    theme(axis.text.x = element_text(size = 8, angle = 45, vjust = 0.5),
          legend.position = "none",
          # axis.text.x = element_blank(),
          # axis.ticks.x = element_blank(),
          axis.title.x = element_blank(),
          strip.background = element_blank(),
          strip.text.x = element_blank(),
          panel.spacing = unit(0, "lines"), #remove space between facets
          panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  pExonic 
  
  pExpr <- exprEnrichCIs %>% 
    mutate(across(matches("SBS"), ~ .x/observed)) %>% 
    pivot_longer(cols = matches("SBS"), names_to = "sig", values_to = "sigCount") %>% 
    #Add dummy lines to replace tissues with no expression data
    bind_rows(tibble(by = rep(c("appendix", "bile_ducts"), each = 4), region = rep(c("Unexp", "0-1", "1-10", "10+"), 2), observed = 100, expected = 1)) %>% 
    mutate(sig = fct_relevel(sig, sigOrder)) %>%
    mutate(region = fct_relevel(region, c("Unexp", "0-1", "1-10", "10+"))) %>% 
    mutate(by = fct_relevel(by, tissueOrder)) %>% 
    ggplot(aes(x= region, y = sigCount, fill = sig, group =1)) +
    geom_hline(yintercept = 1, colour = "grey") +
    geom_bar( stat = "identity") +
    facet_grid(~by, switch = "x") +
    labs(fill = "Sigs") +
    ylab("obs/exp") +
    theme_bw() +
    theme(axis.text.x = element_text(size = 8, angle = 45, vjust = 0.5),
          # axis.text.x = element_blank(),
          # axis.ticks.x = element_blank(),
          axis.title.x = element_blank(),
          strip.background = element_blank(),
          strip.text.x = element_blank(),
          panel.spacing = unit(0, "lines"), #remove space between facets
          panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  pExpr
  
  
  
  ggarrange(pExonic, pExpr, pRep, nrow = 3) 
  
  png(paste0(plotsPath, 'summaryFigSigsProp.png'), height=8, width=12, units = "in", res = 300)
  ggarrange(pExonic, pExpr, pRep, nrow = 3) 
  dev.off()
  
  pdf(paste0(plotsPath, 'summaryFigSigsProp.pdf'), height=8, width=12)
  ggarrange(pExonic, pExpr, pRep, nrow = 3) 
  dev.off()
}
```
